First Season MWA EoR Power Spectrum Results at Redshift 7

### 2016-07-08 ms.tex
JSD comments

The 21 cm hyperfine transition from neutral hydrogen residing in the IGM during reionization 
offers another promising method to study the EoR. Not only is the 21cm signal a 
\emph{direct} probe of the IGM, but, due to the narrow width of the transition and the 
relationship between observed frequency and line-of-sight distance, it can be used to map 
the full three dimensional space of the epoch. (For reviews, see \citealt{Furlanetto:2006}, 
\citealt{Morales:2010}, \citealt{Zaroubi:2013}, and \citealt{Pritchard:2012}) The first generation of instruments with primary science goals 
to detect the highly redshift 21cm signal have been built, including the GMRT (Giant Metrewave 
Radio Telescope, \citealt{Paciga:2013}), LOFAR (LOw Frequency Array\footnote{http://
www.lofar.org}, \citealt{vanHaarlem:2013, Yatawatta:2013}), PAPER (Precision Array for Probing the Epoch of 
Reionization\footnote{http://eor.berkeley.edu}, \citealt{Parsons:2010}), and the MWA 
(Murchison Widefield Array\footnote{http://www.mwatelescope.org}, \citealt{Tingay:2013, Bowman:2013}). 
Due to relatively low signal to noise, these instruments aim for a statistical measurement of 
the EoR in the form of a cosmological power spectrum. 
Complementary to power spectrum experiments are efforts to detect the sky-average global
21-cm signal from the EoR \citep[e.g.][]{Bowman:2010, Sokolowski:2015, Patra:2013, Voytek:2014}.
Meanwhile, the second generation of
21cm EoR interferometers is on its way with the Hydrogen Epoch of Reionization Array (HERA
\footnote{http://reionization.org}, \citealt{DeBoer:2016}), 
and the low frequency Square Kilometer Array (SKA1-Low \citealt{Mellema:2013}),
which will further 
refine the power spectrum measurement in early build-out stages, but will ultimately be 
capable of imaging the ionized bubbles of reionization in its later stages 
\citep{Beardsley:2015,Malloy:2013}.

Using the first generation of instruments, several upper limits have been placed on the 21 
cm EoR signal \citep{Ali:2015, Dillon:2014, Parsons:2014, Jacobs:2015, Paciga:2013}, and 
\citealt{Pober:2015} was able to place constraints on physical reionization models. 
However, much work is to be done to understand the data produced by these arrays. 
Foreground subtraction and isolation has emerged as a priority in the field. Central to most 
analysis strategies is the concept of an ``EoR window" -- the region of Fourier space where 
spectrally smooth foregrounds have been isolated from the isotropic (spherically symmetric 
in Fourier space) cosmological signal \citep{Morales:2006, Bowman:2009}. More recent 
studies have shown the existence of a foreground ``wedge", the result of instrumental mode 
mixing, throwing power from spectrally smooth foregrounds to higher Fourier modes 
\citep{Thyagarajan:2015b, Thyagarajan:2015, Trott:2012, Liu:2014a, Liu:2014b, 
Hazelton:2013, Pober:2013, Thyagarajan:2013, Vedantham:2012, Morales:2012, Datta:2010}. Nevertheless, 
the EoR window is preserved above the wedge, and 
several analysis pipelines are under active development \citep[e.g. B.~J.~Hazelton et al. 2016, in 
prep; D.~A.~Mitchell et al. 2016, in prep; ][]{Jacobs:2016, Trott:2016, 
Dillon:2013, Trott:2014}.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{example_ps.pdf}
\caption{
An example two dimensional, E-W polarization residual power spectrum formed using the 
three hour golden data set
from August 23, 2013. The foreground wedge dominates the region below the solid 
black horizon line, while regions above are mostly noise-like with the exception of horizontal 
coarse band harmonic lines. 
\label{fig:example_ps}
}
\end{center}
\end{figure}

@@

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{2d_res.pdf}
\caption[Deep 2D sub-band power spectra]{
Residual two dimensional power spectra for the three sub-bands used to place limits on the 
cosmological signal. The left panels show the low band, centered at 174.7 MHz, or a 
redshift of 7.1. The middle panels show the mid band, centered at 182.4 MHz, or a redshift 
of 6.8. The right panels show the high band, centered at 190.1 MHz, or a redshift of 6.5. 
The contours show the window used for averaging to one dimension.
\label{fig:2d_ps_subband}
}
\end{center}
\end{figure*}

@@

The resulting one dimensional power spectra are shown in Figure~\ref{fig:1d_ps}, where 
the measured power is shown with solid blue, the 1-$\sigma$ noise level is shown with 
thin red, and the 2-$\sigma$ upper limit for each bin is shown with dashed magenta. 
Where our unbiased estimator is negative, the absolute value is shown, but with a dotted 
blue line. For consistency with the literature, we plot our one dimensional power spectra 
as $\Delta^2(k)=k^3P_{21}(k)/(2\pi^2)$, which has units of mK$^2$.

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{1D_spectra.pdf}
\caption[1D deep power spectra]{
One dimensional power spectra for our three sub-bands and both instrumental 
polarizations. The solid blue line shows the measured power spectrum with step widths 
corresponding to the bin size used in the average. Where the measured signal is negative 
we plot the absolute value with a dotted line. The thin red line is the 1-$\sigma$ noise level, and 
the dashed magenta line is the 2-$\sigma$ upper limit for each $k$ bin. 
A fiducial theoretical model for a fully neutral IGM from \citet{Furlanetto:2006} is shown
in black for reference.
\label{fig:1d_ps}
}
\end{center}
\end{figure*}

@@

Finally, we place our best upper limits in context with the 21cm EoR field as a whole. A 
direct comparison between measurements from different instruments is difficult due to the 
varying methods, redshifts, and scales probed. However, if we approximate the power 
spectrum in units of mK$^2$ to be roughly flat over the scales probed by the MWA, PAPER, 
and GMRT, we can glean a view of the current state of the field. 
The best result from this analysis is an upper limit on the power spectrum of 
$\Delta^2 \le 2.7 \times 10^4$~mK$^2$ at $k=0.27\, h$~Mpc$^{-1}$ and $z=7.1$.
This is a modest improvement over the previous best MWA results \citep{Dillon:2015}, 
which is not surprising given that this integration contains about ten times more data, but 
is systematic limited. Our results are also consistent with the 
40 hour GMRT limit of $\Delta^2\le6.15\times10^4$~mK$^2$ at $k=0.5\, h$~Mpc$^{-1}$ and $z=8.6$ \citep{Paciga:2013}, though 
probe significantly different periods of the EoR. The frontrunner in terms of an upper limit is 
the PAPER experiment \citep{Ali:2015, Jacobs:2015, Parsons:2014}. The PAPER-64 limit of 
$\Delta^2\le502$~mK$^2$ at $z=8.4$ from a 135 day observing campaign is about one and a half orders of magnitude lower 
than our best limit in power units, though again probing a significantly different redshift. 
The MWA analysis is rapidly improving as we uncover and mitigate systematics to realize
the full sensitivity of the telescope.
The red lines in Figure~\ref{fig:1d_ps} indicate the noise level possible 
to reach with these data if systematics can be overcome.

>>>>

The 21 cm hyperfine transition from neutral hydrogen residing in the IGM during reionization 
offers another promising method to study the EoR. Not only is the 21cm signal a 
\emph{direct} probe of the IGM, but, due to the narrow width of the transition and the 
relationship between observed frequency and line-of-sight distance, it can be used to map 
the full three dimensional space of the epoch. (For reviews, see \citealt{Furlanetto:2006}, 
\citealt{Morales:2010}, \citealt{Zaroubi:2013}, \citealt{Loeb:2013}, and \citealt{Pritchard:2012}) The first generation of instruments with primary science goals 
to detect the highly redshift 21cm signal have been built, including the GMRT (Giant Metrewave 
Radio Telescope, \citealt{Paciga:2013}), LOFAR (LOw Frequency Array\footnote{http://
www.lofar.org}, \citealt{vanHaarlem:2013, Yatawatta:2013}), PAPER (Precision Array for Probing the Epoch of 
Reionization\footnote{http://eor.berkeley.edu}, \citealt{Parsons:2010}), and the MWA 
(Murchison Widefield Array\footnote{http://www.mwatelescope.org}, \citealt{Tingay:2013, Bowman:2013}). 
Due to relatively low signal to noise, these instruments aim for a statistical measurement of 
the EoR in the form of a cosmological power spectrum. 
Complementary to power spectrum experiments are efforts to detect the sky-average global
21-cm signal from the EoR \citep[e.g.][]{Bowman:2010, Sokolowski:2015, Patra:2013, Voytek:2014}.
Meanwhile, the second generation of
21cm EoR interferometers is on its way with the Hydrogen Epoch of Reionization Array (HERA
\footnote{http://reionization.org}, \citealt{DeBoer:2016}), 
and the low frequency Square Kilometer Array (SKA1-Low \citealt{Mellema:2013}),
which will further 
refine the power spectrum measurement in early build-out stages, but will ultimately be 
capable of imaging the ionized bubbles of reionization in its later stages 
\citep{Beardsley:2015,Malloy:2013}.

Using the first generation of instruments, several upper limits have been placed on the 21 
cm EoR signal \citep{Ali:2015, Dillon:2014, Parsons:2014, Jacobs:2015, Paciga:2013}, and 
\citealt{Pober:2015} was able to place constraints on physical reionization models. 
However, much work is to be done to understand the data produced by these arrays. 
Foreground subtraction and isolation has emerged as a priority in the field. Central to most 
analysis strategies is the concept of an ``EoR window" -- the region of Fourier space where 
spectrally smooth foregrounds have been isolated from the isotropic (spherically symmetric 
in Fourier space) cosmological signal \citep{Morales:2006, Bowman:2009}. More recent 
studies have shown the existence of a foreground ``wedge", the result of instrumental mode 
mixing, throwing power from spectrally smooth foregrounds to higher Fourier modes 
\citep{Thyagarajan:2015b, Thyagarajan:2015, Trott:2012, Liu:2014a, Liu:2014b, 
Hazelton:2013, Pober:2013, Thyagarajan:2013, Vedantham:2012, Morales:2012, Datta:2010}, 
while others are investigating the spectral structure of point sources themselves \cite[e.g.][]{Offringa:2016}.
The EoR window is still expected to be preserved above the wedge, and 
several analysis pipelines are under active development \citep[e.g. B.~J.~Hazelton et al. 2016, in 
prep; D.~A.~Mitchell et al. 2016, in prep; ][]{Jacobs:2016, Trott:2016, 
Dillon:2015b, Dillon:2013, Trott:2014}.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{example_ps.pdf}
\caption{
An example two dimensional, E-W polarization residual power spectrum formed using the 
three hour golden data set
from August 23, 2013. The foreground wedge dominates the region below the solid 
black horizon line, while regions above are mostly noise-like with the exception of horizontal 
coarse band harmonic lines. 
The vertical streaks at high $k_{||}$ and high $k_{\perp}$ are due to sparse $(u,v)$ sampling
beyond the dense core of the MWA.
\label{fig:example_ps}
}
\end{center}
\end{figure}

@@

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{2d_res.pdf}
\caption[Deep 2D sub-band power spectra]{
Residual two dimensional power spectra for the three sub-bands used to place limits on the 
cosmological signal. The left panels show the low band, centered at 174.7 MHz, or a 
redshift of 7.1. The middle panels show the mid band, centered at 182.4 MHz, or a redshift 
of 6.8. The right panels show the high band, centered at 190.1 MHz, or a redshift of 6.5. 
The top row corresponds to East-West instrumental polarization, while the bottom row 
shows North-South polarization.
The contours show the window used for averaging to one dimension.
\label{fig:2d_ps_subband}
}
\end{center}
\end{figure*}

@@

The resulting one dimensional power spectra are shown in Figure~\ref{fig:1d_ps}, where 
the measured power is shown with solid blue, the 1-$\sigma$ noise level is shown with 
thin red, and the 2-$\sigma$ upper limit for each bin is shown with magenta. 
Where our unbiased estimator is negative, the absolute value is shown, but with a dotted 
blue line. For consistency with the literature, we plot our one dimensional power spectra 
as $\Delta^2(k)=k^3P_{21}(k)/(2\pi^2)$, which has units of mK$^2$.

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{1D_spectra.png}
\caption[1D deep power spectra]{
One dimensional power spectra for our three sub-bands and both instrumental 
polarizations. The solid blue line shows the measured power spectrum with step widths 
corresponding to the bin size used in the average. Where the measured signal is negative 
we plot the absolute value with a dotted line. 
The gray boxes show the $\pm 2\sigma$ error bars on the measured power spectrum.
Where the boxes meet the horizontal axis we are consistent with zero.
The thin red line is the 1-$\sigma$ noise level, and 
the magenta line is the 2-$\sigma$ upper limit for each $k$ bin. 
A fiducial theoretical model for a fully neutral IGM from \citet{Furlanetto:2006} is shown
in black for reference.
\label{fig:1d_ps}
}
\end{center}
\end{figure*}

@@

Finally, we place our best upper limits in context with the 21cm EoR field as a whole. A 
direct comparison between measurements from different instruments is difficult due to the 
varying methods, redshifts, and scales probed. However, if we approximate the power 
spectrum in units of mK$^2$ to be roughly flat over the scales probed by the MWA, PAPER, 
and GMRT, we can glean a view of the current state of the field. 
The best result from this analysis is an upper limit on the power spectrum of 
$\Delta^2 \le 2.7 \times 10^4$~mK$^2$ at $k=0.27\, h$~Mpc$^{-1}$ and $z=7.1$.
This is a modest improvement over the previous best MWA results \citep{Dillon:2015}, 
which is not surprising given that this integration contains about ten times more data, but 
is systematics limited. Our results are also consistent with the 
40 hour GMRT limit of $\Delta^2\le6.15\times10^4$~mK$^2$ at $k=0.5\, h$~Mpc$^{-1}$ and $z=8.6$ \citep{Paciga:2013}, though 
probe significantly different periods of the EoR. The frontrunner in terms of an upper limit is 
the PAPER experiment \citep{Ali:2015, Jacobs:2015, Parsons:2014}. The PAPER-64 limit of 
$\Delta^2\le502$~mK$^2$ at $z=8.4$ from a 135 day observing campaign is about one and a half orders of magnitude lower 
than our best limit in power units, though again probing a significantly different redshift. 
The MWA analysis is rapidly improving as we uncover and mitigate systematics to realize
the full sensitivity of the telescope.
The red lines in Figure~\ref{fig:1d_ps} indicate the noise level possible 
to reach with these data if systematics can be overcome.
